%Acknowledgments

\renewcommand{\baselinestretch}{2}
\small\normalsize
\hbox{\ }
 
\vspace{-.65in}

\begin{center}
\large{Acknowledgments} 
\end{center} 

\vspace{1ex}

Without the help of my colleagues, advisor, friends, family, and my wife, this thesis and the work it embodies would never have happened.  Thanks go to them and to the EXO-200 collaboration, which has provided a supportive and stimulating environment for my work and learning over the past four years.

Particular thanks go to: David Auty, Phil Barbeau, Giorgio Gratta, Steve Herrin, Sam Homiller, Mike Jewell, Tessa Johnson, Tony Johnson, Caio Licciardi, Mike Marino, Dave Moore, Russell Neilson, Igor Ostrovskiy, J.J. Russell, Simon Slutsky, Erica Smith, Kevin O'Sullivan, Tony Waite, Josiah Walton, Liangjian Wen, Liang Yang, Yung-Ruey Yen, and of course my advisor Carter Hall.  All of you have taught me to be the physicist I am today.  Denoising has been a tremendously collaborative effort, and I am grateful that my last project with this collaboration has provided the chance to work with so many of you so closely.

My advisor Carter Hall has given me the freedom and support to try all kinds of crazy things in the EXO analysis, and I am grateful to him for his open-minded insight throughout.  He has also been enormously helpful to me in finishing this dissertation on a very condensed schedule, taking much time from his own life to improve the quality of this work and its presentation; it could not have been what it is without so much of his help and guidance.

Finally, my friends, family, and wife have been extremely supportive of me throughout my time in graduate school.  You have my deepest gratitude for everything you have done over the past five years and before.  Thank you all.




